---
title: "ETC5512: Assignment 4"
author: "Thuy Ngo" 
format: 
  html:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE, 
                      message = FALSE) 
```


```{r}
library(tidyverse)
library(ggplot2)
library(here)
library(sf)
library(readxl)
library(janitor)
```

# Data collection

- Victoria Crime Statistics: <https://www.crimestatistics.vic.gov.au/crime-statistics/latest-victorian-crime-data/download-data>
    - LGA Recorded Offences Year Ending December 2024 
    - LGA Criminal Incidents Year Ending December 2024
- Population: G01 Geopackage
- Annual growth rate of population: <https://www.abs.gov.au/statistics/people/population/national-state-and-territory-population/latest-release#data-downloads>
    - Get population in 2024 to calculate growth rate -> Apply that to census data
- Median rent: G02 Geopackage

# Data processing

## Median rent by suburb

```{r}
geopath_2021_G02 <- here::here("data/raw/Geopackage_2021_G02_VIC_GDA2020/G02_VIC_GDA2020.gpkg")
st_layers(geopath_2021_G02)

median_rent <- read_sf(geopath_2021_G02, layer = "G02_SAL_2021_VIC") |> 
    select(SAL_CODE_2021, SAL_NAME_2021, Median_rent_weekly, geom) |> 
    clean_names()
```

## Population by suburb

 ```{r}
geopath_2021_G01 <- here::here("data/raw/Geopackage_2021_G01_VIC_GDA2020/G01_VIC_GDA2020.gpkg")
population <- read_sf(geopath_2021_G01, layer = "G01_SAL_2021_VIC") |> 
    select(SAL_CODE_2021, SAL_NAME_2021, Tot_P_M, Tot_P_F, Tot_P_P, geom) |> 
    clean_names()
 ```

 ```{r}
vic_pop_2024 <- 7012962
vic_pop_2021 <- 6566092
cencus_pop <- sum(population$tot_p_p)

pop_growth_rate <- (vic_pop_2024/cencus_pop)-1

est_pop <- population |> 
    mutate(across(c("tot_p_m", "tot_p_f", "tot_p_p"), ~round(.x*(1+pop_growth_rate))))
 ```

```{r handling suburb name}
joined_pop_rent <- median_rent |> 
    st_drop_geometry() |> 
    left_join(est_pop, by = c("sal_name_2021", "sal_code_2021")) |> 
    mutate(
    local_government_area = ifelse(
        str_detect(sal_name_2021, " - Vic\\.\\)"),
        str_extract(sal_name_2021, "(?<=\\().+?(?= - Vic\\.)") |> str_to_upper(),
        NA_character_), # Extract LGA only if it contains ' - VIC.)'
    suburb_town_name = str_trim(str_remove(sal_name_2021, "\\s*\\(.*\\)")) |> str_to_upper()) |> # Remove anything inside brackets including the brackets
    select(-sal_name_2021)
```

## Crime statistics

```{r}
incidents <- read_excel("data/raw/Data_Tables_LGA_Criminal_Incidents_Year_Ending_December_2024.xlsx", sheet = "Table 03") |> clean_names()
offences <- read_excel("data/raw/Data_Tables_LGA_Recorded_Offences_Year_Ending_December_2024.xlsx", sheet = "Table 03") |> clean_names()
```

Offence rate = (Offence count/ERP count) *100,000
Criminal incident rate = (Criminal incident count/ERP count) *100,000

```{r sum offence and incidents}
incidents_per_suburb <- incidents |> 
    filter(year == 2024) |> 
    mutate(across(c("suburb_town_name", "local_government_area"), ~str_to_upper(.x))) |> 
    group_by(local_government_area, suburb_town_name) |> 
    summarise(incident_counts = sum(incidents_recorded))

offences_per_suburb <- offences |> 
    filter(year == 2024) |> 
    mutate(across(c("suburb_town_name", "local_government_area"), ~str_to_upper(.x))) |> 
    group_by(local_government_area, suburb_town_name) |> 
    summarise(offence_counts = sum(offence_count))
```


```{r check if unique suburb in census data is actually unique in VCS}
# Get unique combination of LGA and suburb from incidents and offences
lga_suburb <- bind_rows(incidents_per_suburb, offences_per_suburb) |> 
    distinct(local_government_area, suburb_town_name)

# Get list of suburb from median rent
suburb_list <- joined_pop_rent |> 
    distinct(suburb_town_name) |> 
    pull(suburb_town_name)

# Count how many times each suburb appear in lga_suburb
count_suburb <- lga_suburb |> 
    count(local_government_area, suburb_town_name) |> 
    rename(count = n) |> 
    left_join(lga_suburb, by = c("local_government_area", "suburb_town_name")) |> 
    mutate(count = ifelse(is.na(count), 0, count))
```


```{r}
# Separate joined_pop_rent into two dataframes of NA lga and non-NA lga
unique_suburb <- joined_pop_rent |> 
    filter(is.na(local_government_area))

non_unique_suburb <- joined_pop_rent |> 
    filter(!is.na(local_government_area))

# Join non_unique_suburb with incidents_per_suburb
non_unique_suburb_merged <- non_unique_suburb |>
    left_join(incidents_per_suburb, by = c("suburb_town_name", "local_government_area")) |> 
    left_join(offences_per_suburb, by = c("suburb_town_name", "local_government_area"))

# Join unique_suburb with incidents_per_suburb
unique_suburb_merged <- unique_suburb |>
    left_join(incidents_per_suburb, by = c("suburb_town_name")) |> 
    left_join(offences_per_suburb, by = c("suburb_town_name", "local_government_area.y" = "local_government_area"))
```


```{r find duplicated suburb in unique_suburb_merged}
# Find duplicated suburb in unique_suburb_merged
duplicated_suburb <- unique_suburb_merged |>
    group_by(suburb_town_name) |>
    summarise(count = n()) |> 
    filter(count > 1) |> 
    pull(suburb_town_name)

unique_suburb_merged |>
    filter(suburb_town_name %in% duplicated_suburb) |> 
    View()
```


```{r sum incidents and offences for duplicated suburb}
# all duplicated suburb in unique_suburb_merged extend to more than 1 LGA
# so we need to sum incidents and offences for each LGA

group_col <- c("sal_code_2021", "median_rent_weekly", "tot_p_m", "tot_p_f", "tot_p_p", "geom", "local_government_area", "suburb_town_name")

unique_suburb_merged |> 
  select(-local_government_area.x) |>
  rename(local_government_area = local_government_area.y) |>
  group_by(across(all_of(group_col))) |>
  summarise(
    incidents = sum(incident_counts, na.rm = TRUE),
    offences = sum(offence_counts, na.rm = TRUE),
    .groups = "drop") |> 
    bind_rows(non_unique_suburb_merged)
```